---
title: |
    Whose voices are heard? [^1]

    [^1]: **Draft version. Please do not circulate or cite without permission.**
subtitle: "Making sense of framing patterns in protest coverage"
format:
  aft-pdf:
    include-in-header:
      - text: |
          \usepackage{siunitx}
          \usepackage{graphicx}
          \usepackage{rotating}
  aft-html: default
author:
  - name: Johannes B. Gruber
    affiliations:
      - name: Vrije Universiteit Amsterdam
        department: Department of Communication Science
    orcid: 0000-0001-9177-1772
    email: j.b.gruber@vu.nl
    url: https://www.johannesbgruber.eu/
date: today
filters:
  - abstract-section
keywords: [protest paradigm, media framing, social movements, news reporting, content analysis]
reference-section-title: References
bibliography: references.bib
execute:
  echo: false
  eval: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.path = "plots/"
)
source("../misc/functions.R")
library(tidyverse); theme_set(theme_minimal())
library(kableExtra)
library(texreg)
library(margins)
library(grid)
```

# Abstract

```{r article, child = "abstract.md"}
```

# Introduction

Protest is a fundamental democratic tool for citizens to voice dissent about issues in society.
In modern democracies, however, citizens usually only hear about protest through mass media.
And this hearing consists not only of the audience becoming aware of a protest event, but ideally should also involve them receiving the information to understand the underlying issue and the causes for the public dissent.

> Which factors explain the choice of frames by the news media when covering protest events?

While the concept of testing factors which condition protest coverage is not new [e.g., @boyleAdherenceProtestParadigm2012; @kilgoHierarchy2019; @leeTriggeringProtestParadigm2014], the set of variables presented in this section is novel and, to my knowledge, the most complete so far.

# Protest Coverage in News Media

## The Protest Paradigm

Most of the literature concerning protest coverage in news media uses the *protest paradigm* concept to explain a recurring pattern in news about demonstrations:
the media focus on crime, clashes with police and the spectacle surrounding a protest or the protesters themselves and their looks, demographics or occupation, while neglecting to explain *why* protesters spent time and resources on organising and participating in a street demonstration in the first place.
The *protest paradigm* explains this by presuming a mental catalogue, supposedly common among journalists, which indicates to them "where to look (and where not to look), and informs them about what to discover" [@chanJournalisticParadigmsCivil1984, p. 187].
The *protest paradigm* is thus related to the literature on the sociology of newsmaking, which observed that news workers are "routinizing the unexpected" [@tuchmanMakingNewsDoing1973, p. 110] by creating practical and effective procedural rules, routines and mental catalogues of news story themes to cope with the onslaught of new events.

Most of the focus in previous studies has been spent on the dangers of this routinised reporting.
This is due to the normative core of the *protest paradigm*:
by regularly focusing on the action itself, rather than the sometimes more elusive and complicated message [@smithProtestAgendaBuilding2001] the *protest paradigm* could lead to delegitimisation and marginalisation of protest events, protest movements, or protest as a political resource [@mcleodSocialControlSocial1999].

Evidence for the *protest paradigm* has been found in most studies that analysed news coverage of protests over the years [for a systematic overview see @gruberTrouble2023].
However, more recent work has also asked *how common* delegitimising devices of the *protest paradigm* are.
While it seems fairly clear that *some* coverage follows routinized dismissive patterns, this might not be that damaging to protests as long as substantive information about causes and grievances are still available elsewhere.
Studies following this logic suggest that the delegitimising aspects of the *protest paradigm* might occur in reporting more conditionally [@kilgoHierarchy2019; @woutersReportingDemonstrationsEpisodic2015], and less frequently over time [@gruberTrouble2023; @araizaHandsDontShoot2016; @kyriakidouIndignadosProtestsSpanish2017].

## What Drives Protest Coverage? {#sec:factors}

This moves the conditions under which the delegitimising aspects of the *protest paradigm* are applied into focus.
How these factors might shape reporting in practice is provided, for example, by @tenenboimweinblattProducingProtestNews2014, who studied in in-depth interviews how journalists perceived their hand in framing a social protest in Israel.
At first, individual level factors, namely the strong indetification of journalists with the sociodemographically protesters, who belonged to a similar socio-economic class, and their message, which criticised the same issues reporters faced in their daily lives, resulted in a sympathetic stance and more supportive coverage.
Yet, journalists noted that this identification and sympathy for the protest demands collided with factors such as professional and procedural norms, which kept their reporting "objective" with less supportive coverage than some had hoped for.
As events progressed, substantial economic backlash from investors and advertisers, who felt targeted by protests, turned the table.
Suddenly economic and organisational demands became more noticeable to journalists.
In effect, journalists ceased to portray the protests in a favourable light after initial support.^[see also @shultzinerjournalistsprotest2018 and @shoemakerMediatingMessage21st2014 for a more general theory on factors that shape news.]
<!-- This is in line with @curranPowerResponsibilityPress2007, who have argued that the historical control of the press by political actors has been replaced over time by market forces, which they believe to be even more effective in eliminating radical ideas from public discourse. -->

What this example makes clear is that there is rarely just one factor that shapes reporting and that in order to explain reporting, the range of included factors should be as complete as possible.
The literature has identified several of these factors that can be broadly divided in event-level, outlet-level and time-related factors.[^Similar to how @shoemakerMediatingMessage21st2014 divide influences on factors believed to shape the news into levels ordered from micro to macro.]
Most of these studies measure a dividie between *delegitimising* and *legitimising* protest coverage.
Although this dichotomy is a little simplistic [@harlowNewParadigm2023] it serves the purpose of formulating theoretical expectations here and to make it easier to discuss previous findings.

<!-- Note that at this stage, it is unknown which frames the analysis will reveal, making it impossible to specify which frames might correlate with which factors.
Many previous studies worked with the dichotomy of coverage adhering to the *protest paradigm* and coverage that is more supportive of a protest goal --- which I will call *delegitimising* and *legitimising* coverage in the following.
As it is expected that most frames can be placed within this broad dichotomy, the hypotheses in this section capture the expected relation between factors and a potential set of frames that are, most likely, legitimising or delegitimising.
In other words, while the analysis allows the revelation of more fine-grained frames, this section will discuss the theoretical expectations for relationships between the factors with legitimising and/or delegitimising coverage of protest.
However, the dichotomy of legitimising and delegitimising coverage is unlikely to tell the entire story.
What aspects of an event do news highlight and which do they omit specifically?
Depending on the specific framing, the degree to which reports legitimise or delegitimise a protest cause could vary substantially.
So while the described dichotomy is useful to discuss factors that drive reporting, the ultimate goal is to measure the influence these factors might have on the usage of specific frames. -->

### Event-Level Factors

#### Protest Goals

On the level of the individual event, several studies document that the *goals or causes* voiced by protesters shape the coverage these events receive.
In earlier studies, this has been mostly attributed to the idea that the media are agents of social control, who consciously and unconsciously reward conformity to and punish deviance from a perceived mainstream or *status quo* [@mcleodSocialControlSocial1999].
In this regard, @shoemakerMediaTreatmentDeviant1984 found that groups that are perceived as more deviant or radical tend to receive more delegitimising coverage [@hallPolicingCrisisMugging1978]^[Also see the deviance sphere in @hallinUncensoredWarMedia1986.].
Based on this, @boyleInfluenceLevelDeviance2004 tested how the *level of deviance* of protests shape their coverage.
They conclude that coverage of deviant protest conformed closer to the *protest paradigm*, conceptualised as more critical coverage, coverage employing episodic framing and coverage that is less likely to use protesters as sources.

The idea of a *level of deviance* can also be linked to @bennettindexing1990's theory of "indexing", which postulates that the range of voices and interpretations of events are organised "according to the range of views expressed in mainstream government debate about a given topic" (p. 106).
This means that in times of elite consensus, media will limit its discussion to the opinions expressed by the powerful.
Only in times of elite dissensus, the voices of the broader public will be heard.
Otherwise, alternative information or narratives are assumed to encounter fierce resistance or outright blockage [@bennettNoneDareCall2006].
Generally, it is thought that indexing hurts protesters' ability to explain the causes of their efforts [@andrewsMakingNewsMovement2010].

However, while the notion that deviant goals are less likely to be cast in a positive light is well supported, newer studies have departed from it due to a theoretical ambiguity:
there is no irrefutable conceptualisation of the "status quo" as political and social realities are constantly changing [@kilgoHierarchy2019; @harlowNewParadigm2023].
A goal that was a threat to the status quo five or ten years ago could be in favour of the current situation.
<!-- @boyleMeasuringLevelDeviance2009, for example, tested this idea by comparing the treatment of pro-life and pro-choice protests before and after the ruling in the Roe v. Wade case --- which effectively legalised abortion in the entire US.
They found only partial support for the importance of the status quo for protest coverage as pro-life protesters were treated more critically after their goal became to change the status quo, but pro-choice protesters were treated the same before and after the law changed.
Similarly, @boyleNewspapersProtestExamination2005 found that while the level of deviance of most protest movements decreased over time, headline and article valence stayed roughly the same. -->

Analysing the deviance of protest goals from the status quo, therefore, appears less meaningful and has shifted the attention to more clear cut categories.
@boyleInfluenceLevelDeviance2004 and @boyleNewspapersProtestExamination2005 studied a range of events between 1960 and 1999 and found that anti-war protesters were treated most critically, followed by labour protests (e.g., strikes and pickets), protests against wrongdoings by the police [only @boyleInfluenceLevelDeviance2004] and social-issue protests (e.g., abortion or hunger) which were treated most favourably.
In an analysis of newer coverage (2007-2009), @boyleAdherenceProtestParadigm2012 found that *political protests*, by which they mean protests seeking political change, were treated more critically than anti-war and social-issue protests.

Instead of ranking these types of protest according to the level of deviance, @kilgoHierarchy2019 suggest that "journalistic routines are subject to a *hierarchy of social struggle*, in which certain topics are given precedence and legitimacy, and others are delegitimised, trivialised, or ignored altogether" (p. 16).
They also add several new types to this list.
Specifically, they found that coverage of protest against anti-Black and anti-Indigenous racism, international protest, and anti-Trump protests use frames that delegitimise their issue far more often than other protest types.
For some issues, like health, the environment and immigration protests, the press used legitimising frames more often.
Similarly, @harlowWholeWorldWatching2020 coded 33 specific types of protests before summarising them into six categories.
They find that while not all protest types had a significant influence on the use of frames, some patterns emerged:
conservative protests, for example, were less likely to be covered in a riot frame but more likely to be covered in a spectacle frame, while socio-economic protests were more likely to be covered using the riot frame.
For a legitimising frame, dubbed debate frame, no significant relationship to a protest type was found.

Following @kilgoHierarchy2019, I expect that some goals are more likely to receive legitimising and some are more likely to receive delegitimising coverage.
In other words, I expect that a *hierarchy of social struggle* will emerge in UK reporting of protest, which can predict which causes are given precedence over others.
I will use five broad categories for protest goals, ordered from most to least likely to receive critical coverage these are:
*political*, *anti-war*, *labour*, *police* and *social-issue* protests [@boyleAdherenceProtestParadigm2012; @boyleNewspapersProtestExamination2005; @boyleInfluenceLevelDeviance2004].
The hypotheses I will test are thus:

- `r hyp("H1a")`
- `r hyp("H1b")`

Note that the expected hierarchy in **H1b** is the flipped version of **H1a** as I expect that the likelihood for a protest to be covered with a legitimising frame should be roughly diametrically opposed to the likelihood for a protest to be covered with a delegitimising frame.

#### Tactics

Early studies often treated goals and tactics as two sides of the same coin.
However, @boyleMeasuringLevelDeviance2009 established that this is a shortcoming as both have distinct effects on the coverage of protest.
Specifically, they showed that when both are taken into account, extreme protest tactics facilitate critical coverage stronger than protest goals.
This insight had already been observed by one of the earliest studies of protest coverage:
@gitlinWholeWorldWatching1980 found that the amount of outright negative coverage, instead of trivialisation, increased as anti-war protests became more extreme.
@boyleAdherenceProtestParadigm2012, who additionally test the effects of protest location, protest type, and a group's goals in an international comparison of protest coverage, conclude that the tactics employed during a protest are the strongest determinant of delegitimising coverage.
Likewise, @WasowSeeding2020 concludes that nonviolent black-led protests during the civil rights movement led to more coverage of civil rights issues, while protester violence invoked language associated with disorder and calls for law and order.

How extreme the tactics are is usually determined by two variables:
do the protesters break laws and do they engage in violence?
From a broader perspective, a focus on extreme protest tactics is, therefore, in line with the professional norms of "newsworthiness" mentioned above:
usually, the clashes rather than the message of a protest will produce most conflict, drama, unusualness and negativity.
This leads to the following expectations:

- `r hyp("H2a")`
- `r hyp("H2b")`

The relation between violent tactics and coverage has been known for a long time by both researchers and protesters [@gitlinWholeWorldWatching1980].
When tactics become more violent, coverage becomes less supportive of the cause and more supportive of repressive measures.
Nevertheless, protesters often use violent tactics because this will usually lead to (more) coverage [@boyleAdherenceProtestParadigm2012].
**H2a** and **H2b**, therefore, also answer the question if this bargain makes sense for a movement.


#### State Response

As mentioned above, @WasowSeeding2020 found in his study of black-led protests between 1960 and 1972 that protester-initiated violence led to delegitimising coverage and subsequently turned public opinion and voter preferences against the civil rights movement.
However, @WasowSeeding2020 also tested the effects of violent state response towards the movement.
Specifically, he found that there are four scenarios that yielded slightly different outcomes:
When the state tolerated peaceful protest, this lead to coverage which legitimised the concerns voiced by protesters;
when the state used violent repression against peaceful protesters, this lead to coverage that is more supportive of the protesters' goal --- and to an overall higher volume of coverage about the respective event;
when only the protesters were violent, the civil rights movement was penalised in the coverage;
the same was, however, true when both police and protesters engaged in substantial violence.
In this last case, the coverage is predicted to focus on concerns about order and public safety.

Wasow's [-@WasowSeeding2020] study thus mostly confirms the relationship between violence and (de)legitimising coverage formalised in **H2a** and **H2b**.
A scenario in which the state uses violent repression against peaceful protesters, however, has not been covered by any previous studies.
Following @WasowSeeding2020, I include this additional layer, which leads to the expectation that:

- `r hyp("H3")`

### Outlet-Level Factors

#### Newspaper Ideology

In the landmark study which defined the idea of a *protest paradigm*, @chanJournalisticParadigmsCivil1984 found that not all newspapers engage in delegitimising protests in the same way:
the left-leaning media were less hostile towards the protesting teachers and students in Hong Kong than centrist media, which were in turn less hostile than right-leaning journalists.
This factor has usually been overlooked in subsequent US-based studies, which might have assumed that ideological differences between major outlets are less important given the emphasis on objectivity in the country's journalistic culture.
More recent studies, however, have started again to employ newspaper ideology as an explaining factor for the level of delegitimising reporting different protests face [@leeTriggeringProtestParadigm2014; @shahinProtestingParadigmComparative2016; @ozMainstreamMediasCoverage2016].
UK media --- the focus of this study --- are known to be distinctly more partisan than outlets in the US and elsewhere [@kuhnPoliticsMediaBritain2007; @hallinComparingMediaSystems2004], which is likely to increase the effect of newspaper ideology on protest reporting.
This leads to the expectation:

- `r hyp("H4a")`
- `r hyp("H4b")`

@weaverRevisitingProtestParadigm2012 add nuance to this:
more than by their right or left ideology, the coverage of an outlet might be determined by the distance of their own ideological tilt to the ideology of a protest in question.
They test this using prime-time cable news coverage about the right-leaning *Tea Party* movement in the United States in 2009-2010 to assess which outlets used marginalisation devices more often.
Their finding suggests that liberally aligned MSNBC was most likely to marginalise the *Tea Party*, followed by the centrist AP and CNN, and Fox News being supportive overall.
Since protests have traditionally held goals that were considered left more often in Western countries --- and continue to do so [@borbathDifferentWorldsContention2020] --- the assumption that right-leaning media are generally more critical of protest might be incorrect.
This leads to two hypotheses that will be tested:

- `r hyp("H5a")`
- `r hyp("H5b")`

#### Type of Newspaper

Gavin [-@gavinPressTelevisionBritish2007, pp. 95-118] and @rosieAnarchistsWorldCup2009 add another factor related to newspaper outlets:
the difference between tabloid and broadsheet newspapers.
In general, it is expected that tabloid newspapers are more focused on human interest and entertainment [@kuhnPoliticsMediaBritain2007; @hallinComparingMediaSystems2004].
The sharp divide in reporting style between broadsheets newspapers (*The Daily Telegraph*, the *Financial Times*, *The Guardian*, *The Independent* and *The Times*), and tabloids (*Daily Express*, *Daily Mail*, the *Daily Mirror* and *The Sun*) is a well-known factor in the UK media system and should hence not be neglected.

@gavinPressTelevisionBritish2007, who assessed the May Day demonstration in London in 2000, concluded that tabloid outlets used the delegitimising elements from the *protest paradigm* more extensively and failed to mention any motivation or goals of the protest.
Broadsheet newspapers --- especially *The Guardian* --- apparently made an effort to explain why the protests were held.
In their qualitative study, @rosieAnarchistsWorldCup2009 find that tabloid newspapers were more extreme in their framing of peaceful G8 protesters as a fringe group of radical militants.
It is, therefore, only plausible to expect that the two types use different frames.
Following the work by @rosieAnarchistsWorldCup2009, the following expectations can be made:

- `r hyp("H6a")`
- `r hyp("H6b")`

### Time-Bound Factors
#### Year of Protest

Two factors related to time are expected to have an influence on reporting about protests.
The first one is that protests that happened more recently receive more legitimising coverage.
In an earlier study, I showed that the level of legitimising framing in news about domestic protest has been steadily growing since the early 1990s [@gruberTrouble2023].
Several mechanisms have been suggested for this:
Protest has largely been normalised since the first seminal studies about movements in the 1960s have been conducted[@daltonDemocraticChallengesDemocratic2004; @sandersDynamicsProtestBritain2003; @norrisDemocraticPhoenixReinventing2002, pp. 188-212; @vanaelstWhoThatWo2001; @oliverPoliticalProcessesLocal2000].
Today, many different groups employ protest as a tool, and many more people today have participated in a protest at least once in their lives [e.g., @borbathDifferentWorldsContention2020; @HutterChallenges2019].
Besides this, information about protesters' goals and grievances are more readily available on the internet [@humprechtMoreSameMarketplace2013; @posterWhatMatterInternet2001; @boydSocialNetworkSites2011], while journalists likely feel more pressure to paint a complete picture today as audiences have a sharply increased pool of alternative information sources [@vanaelstPoliticalCommunicationHighchoice2017].

Whethter it is a combination of mechanisms or other unknown forces, the expected result of this factor is:

- `r hyp("H8a")`

#### Days Since Protest

The second factor very likely only affects large scale, highly mediatised protests.
@gottliebProtestNewsFraming2015 suggests that coverage follows a *news framing cycle*:
coverage of protest begins with a focus on conflict --- hence marginalising or ignoring the goals and grievances of a protest --- then shifts to more substantial coverage before returning to conflict.
This concept, based on @downsAttentionCycle1972, follows the logic that journalists will use different frames to cover an event as time passes "to keep the story alive and fresh" [@chyiMediaSalienceProcess2004, p. 22].
<!-- Also see MouraoFerguson2018 -->

The *news framing cycle* intuitively makes sense, yet the proposed time frame might suffer from one problem:
Gottlieb’s [-@gottliebProtestNewsFraming2015] case, the Occupy Wall Street movement, was arguably special since protest stretched over a long period and media attention did not wane for months.
He, therefore, suggests that the second phase, when reporting shifts to more substantial coverage, started only in the third week after the protest started.
This finding is probably not suitable for generalisation as attention to most protests wanes after days, rather than weeks or months.
In most cases, the media will have forgotten about a protest the day after the first reports were printed.

Nevertheless, the idea of a protest news framing cycle seems worth testing, albeit in an abbreviated form:

- `r hyp("H8a")`
- `r hyp("H8b")`

# Data and Methods

To test these hypotheses, I combine two datasets.
The first was created through content analysis and contains measures of framing in reports about a large number of events in UK mainstream newspapers from 1992 to the end of 2017.
The second one contains information about protest events.
These data are matched and the hypotheses are tested through multilevel logistic regression models.
@tbl-variables shows the dependant and independent variables of these models --- one for the presence or absence of each frame.
How this data was sourced is explained below.

```{r variables}
#| label: tbl-variables
#| tbl-cap: Model Variables
variables <- rio::import("../misc/variables.csv")  |>
  mutate(across(everything(), ~gsub('"', '', .x, fixed = TRUE)))  |>
  select(-`Levels-long`, -Mean, -Category)

if (isTRUE(knitr:::pandoc_to() == "latex")) {
  variables  |>
    custom_df_print(caption = "Model Variables",
                    font_size = 9,
                    footnote = "Reference category",
                    latex_options = "scale_down")  |>
    pack_rows("Dependent", 1, 1)  |>
    pack_rows("Event-Level", 2, 4)  |>
    pack_rows("Outlet-Level", 5, 7)  |>
    pack_rows("Time-Bound", 8, 9)  |>
    pack_rows("Level 2", 10, 10)  |>
    column_spec(5, width = "3.7cm")
} else {
  variables  |>
    custom_df_print(caption = "Model Variables", font_size = 9)
}
```

## Content Analysis

I use media coverage of protests as the dependent variable.
Newspaper articles were employed instead of alternatives, such as online news, social media content or TV transcripts for two reasons.
Firstly, despite declining audiences, mainstream news media such as newspapers are still deemed the most important platforms for public debate and the creation of the public agenda [@rogstadTwitterJustRehashing2016].
Newspapers might have lost some of their intermedia agenda-setting power due to being slower to release breaking news, but the more in-depth writing produced by newspapers nevertheless forms a crucial share of what is redistributed via online news and social media and is key in initiating, amplifying and sustaining attention to an issue [@Langerwindrush2020; @harderIntermediaAgendaSetting2017; @rogstadTwitterJustRehashing2016].
Secondly, suitable data from UK newspapers were already archived and digitally available for the whole period of interest.

@tbl-outlets presents the selection of newspaper outlets, strategically chosen to contain variance on the two important factors:
ideology and type of newspaper in the UK
Additionally the selection was made to include the outlets with the highest circulation figures.
I retrieved articles from LexisNexis, utilizing the keywords "protest" and "demonstration" and several misspellings.
A pilot ensured the comprehensiveness of these keywords by testing a range of synonyms like "march", "rally", or "direct action", which, however, did not yield any additional relevant articles.
The articles underwent a processing phase using the R-package LexisNexisTools [@gruberlnt2021] which consisted of removing duplicate entries, items discussing non-domestic protests, and false positives.
To maintain theoretical integrity, I opted to limit our data to domestic protests, as studies have indicated notable differences in the coverage of non-domestic protests [@boyleAdherenceProtestParadigm2012].
This led to the final dataset, comprising 27,496 newspaper articles.

```{r outlets}
#| label: tbl-outlets
#| tbl-cap: Selection of Newspaper Outlets for Data Collection
outlets <- tibble::tribble(
  ~Type,        ~Ideology,  ~Outlet,                 ~"Available from", ~"Circulation (2005 -- 2017)",
  "Tabloid",    "Left",    "The Mirror",             1995,              "1,720,000 -- 692,295",
  "Tabloid",    "Right",   "The Sun",                2000,              "3,273,000 -- 1,602,320",
  "Tabloid",    "Right",   "Daily Mail",             1992,              "2,426,000 -- 1,442,924",
  "Broadsheet", "Left",    "The Guardian",           1992,              "366,645 -- 153,431",
  "Broadsheet", "Left",    "The Independent",        1992,              "263,595 -- NA",
  "Broadsheet", "Right",   "Financial Times",        1992,              "419,386 -- 190,046",
  "Broadsheet", "Right",   "Daily Telegraph",        1992,              "460,585 -- 907,329",
  "Broadsheet", "Right",   "The Times",              1992,              "679,190 -- 440,736"
)

footnotes = c(
  "Source for circulation numbers: http://www.magforum.com/papers/nationals.htm",
  "No figures available for The Independent in 2017 as the outlet moved to online-only publication in 2016."
)
library(gt)
outlets |>
  gt() |>
  tab_footnote(
    footnote = "Source for circulation numbers: http://www.magforum.com/papers/nationals.htm",
    locations = cells_column_labels(columns = "Circulation (2005 -- 2017)")
  ) |>
  tab_footnote(
    footnote = "No figures available for The Independent in 2017 as the outlet moved to online-only publication in 2016.",
    locations = cells_body(columns = "Circulation (2005 -- 2017)", rows = 5L)
  )
```

This dataset was coded for framing of protest reporting.
Specifically, I re-use the measurement of framing in newspaper reports on protests and demonstrations from @gruberTrouble2023.
In short, frames were identified by coding the frame elements found in Entman's [-@entman1993] definition of framing (problem definition, causal interpretation, moral evaluation, and treatment recommendation) in 500 randomly selected articles and combining the coded elements to frames using factor analysis.
This results in one binary variable per frame, which takes the value 1 if a frame is present and 0 otherwise.
Each article can contain several frames, or none.
After training supervised machine learning models trained on the manually coded random sample, all remaining articles were coded using these models.

@fig-frames show the seven identified frames and how often they were coded in news articles.
The majority of them were determined to delegitimise protest using many of the same patterns identified in the *protest paradigm* literature.
They portray protesters as a fringe group of freaks who disturb the social norms(Decay of Morals), highlight the violence caused by protesters (Troublemakers), suggest that protest is a bothersome interruption of everyday activities (Nuisance), or focus on police actions to reinstate public order (Law & Order).
Two legitimising frames, on the other hand, describe the main topic and cause of a protest event (Cause & Grievances) or highlight the accomplishments of the protesters for a just cause (Righteous Struggle).
The remaining frame, Police Violence can neither be said to be legitimising nor delegitimising, highlight the shortcomings of this dichotomy [@harlowNewParadigm2023].
The frame focuses on violence, which is usually assumed to distract from protest demands and hence delegitimising.
Contrary to most theoretical expectations, however, the frame portrays police as responsible for violence.
This could lead to more interest in and sympathy towards a protest goal.

```{r}
#| label: fig-frames
#| fig-cap: Number of Articles in Which Each Frame Is Present
readRDS("../misc/frame_distribution.rds") %>%
  ggplot(aes(y = frame, x = n, fill = super)) +
  geom_col() +
  scale_fill_manual(values = jbg_pal(sel = c("darkred", "darkblue", "black"))) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

<!-- In its most advanced, and probably broadest, conceptualisation, @mcleodSocialControlSocial1999 state that the *protest paradigm* combines *unfavourable story framing*, *reliance on official sources and official definitions* and the *invocation of public opinion* (e.g., the protesters are a minority). -->
<!-- "Protest coverage research needs to move beyond the dichotomous positive versus negative framing to account for the ways in which violence can in fact be a legit- imate way to attract media attention and legitimize protesters’ grievances and tactics (Cammaerts 2013; Drury and Reicher 2018; Saavedra and Drury 2019)" -->

## Independent Variables

This article tests the relation between framing of protest and eight independent variable identified from the literature above:
*Protest Goal*, *Protester Violence*, *State Repression*, *Newspaper Ideology*, *Newspaper Type*, *Ideological Divide*, *Days Since Start*, and *Year of Protest*.
*Newspaper Ideology* were obtained through the meta information of the coded news articles *Newspaper Type*.
For all other variables the *Mass Mobilization Project*'s [@ClarkMMPdata2019] dataset on protest events was employed.
This dataset was chosen for practical reasons as it contains the variables necessary and covers the chosen time frame (1992-2017).
@ClarkMMPdata2019 use the keywords "Protest", "Demonstration", "Riot", and "Mass Mobilization" on the *LexisNexis* newspaper archive, selecting "All News" as the source.
For the selected time frame in the UK, the *Mass Mobilization Project* lists 471 events between 1992 and the end of 2017, as shown in @fig-mmpyear.

```{r}
#| label: fig-mmpyear
#| fig-cap: "Number of Protest Events in the UK per Year per Clark and Regan (2019)"
readRDS("../misc/protest_event_data_uk.Rds")  |>
  mutate(year = lubridate::floor_date(date_start, "year"))  |>
  group_by(year)  |>
  count()  |>
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = jbg_pal(sel = "grey")) +
  scale_x_date(labels = lubridate::year, breaks = "year",
               expand = c(0, 0)) +
  #scale_y_continuous(labels = scales::percent, limits = c(0, 0.7)) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank())
```

To merge the two datasets, I employed a two-staged strategy:
first, I selected all articles in the newspaper dataset that were published no earlier than seven days before an event started and no later than 30 days after it ended;
then I used the variables "Protest Locations", "Protest Group Identity" and "Protester Demands" to extract suitable keywords [see Appendix].
I then used keywords to filter the articles selected in the first stage:
articles were discarded if they did not feature at least one of the keywords.
This selection was manually validated and overlaps between events were separated by hand.
Overall, 5,639 articles could be matched with the 471 events from the database.
Most independent variables in the raw data were either enriched or recoded before the main analysis [see Appendix].

Using the variables listed in @tbl-variables, I fitted multilevel logistic regression models (one for each frame) to establish the influence of the different factors on reporting.
Specifically, I used the *lme4* package in R [@lme4].
This is appropriate as the dependent variables for all models are binary and the articles published in the same newspaper and in the same year effectively constitute a panel dataset since observations must be seen as nested and not independent.
It was considered to use newspapers as the second level variable as unobserved and unobservable heterogeneity is most likely present between outlets.
However, the number of newspapers in the dataset (eight) is considered problematically low for fitting a valid multilevel model [e.g., @bryanmultilevel2016].
Therefore, I used newspaper-year as the Level 2 variable ($n = 180$).
<!-- That means that *The Guardian*, for example, is treated as a different second-level case in 1992, 1993, 1994 and so on.
This makes sense as the last chapter has shown that there is a difference in reporting between all newspaper over time. -->
I assessed whether adding the newspaper-year level increases or decreases model fit.
To compare models, the Akaike information criterion (AIC) and the Bayesian information criterion (BIC) are calculated and compared using Analysis Of Variance (ANOVA) tests.
The lower the AIC and BIC, the better the model fits the data.
Both AIC and BIC are always considerably lower when including the second level into the modelling.
The p-values suggest that including the random effect improves all models significantly.
I examined if adding a newspaper-year level improves model fit.
The model comparisons were done using AIC, BIC, and ANOVA tests.
The inclusion of the second level in modelling significantly lowers both AIC and BIC, and the p-values indicate a significant improvement in all models.

# Results

Discussion of results is divided into two steps.
I first refer to the frames aggregated into legitimising and delegitimising to test the hypotheses set up above.
This, however, neglects some of the details revealed through the individual seven frames.

## (De)Legitimising Framing

```{r}
#| results: asis
models <- readRDS("../misc/models.Rds")
model_logit_dele_mlm <- models$model_logit_dele_mlm
model_logit_le_mlm <- models$model_logit_le_mlm

if (is_latex()) {
  texreg(
    c(model_logit_dele_mlm, model_logit_le_mlm),
    caption = "Regression Results for Aggregated Frames",
    custom.model.names = c("Delegitimising Frames", "Legitimising Frames"),
    custom.coef.names = c(
      "(Intercept)",
      "goal: anti-war",
      "goal: labour protests",
      "goal: police",
      "goal: social-issue",
      "violent protest",
      "repression of peaceful p.",
      "right-wing",
      "tabloid newspaper",
      "ideological divide: conflict",
      "ideological divide: ambiguous",
      "days from start",
      "year of protest"
    ),
    groups = list("Event-Level Factors" = 2:7, "Outlet-Level Factors" = 8:11, "Time Bound Factors" = 12:13),
    digits = 3,
    stars = c(0.001, 0.01, 0.05, 0.1),
    symbol = "+",
    #scalebox = 1,
    siunitx = TRUE,
    single.row = FALSE, # True looks better here but much worse on the other. To keep it consistent
    booktabs = TRUE,
    label = "tab-regtwoframes",
    use.packages = FALSE
  )
} else {
  data.frame(placeholder = "In the PDF, the table is around here") %>%
    custom_df_print(caption = "Regression Results for Aggregated Frames")
}
```

```{r}
model_margins <- readRDS("../misc/model_margins.Rds")
twofram_marg <- list(model_margins$model_logit_dele_mlm,
                     model_margins$model_logit_le_mlm) %>%
  lapply(function(m) {
    summary(m) %>%
      as_tibble() %>%
      filter(p < 0.05) %>%
      mutate(model = colnames(m)[1])
  }) %>%
  bind_rows() %>%
  mutate(factor = clear_var_names(factor))

marg <- function(...) {
  marg_default(df = twofram_marg, ...)
}
```

Table 3 shows the models for the legitimising and delegitimising frame aggregates.
Since raw values of coefficient estimates are less useful in a logisitc regression, I additionally I additionally provide Average Marginal Effects (AME) in Figure @fig-margtwoframes.
Holding all other covariate values constant, AMEs express here by how much the probability of using a frame increases or decreases *on average*, if a dependent variable changes by 1 in their respective units.
AMEs were calculated using the *margins* package in R [@marginspackage].

```{r}
#| label: fig-margtwoframes
#| fig-cap: Average Marginal Effects for Significant Variables in Aggregated Frame Models
plot <- twofram_marg %>%
  ggplot(aes(x = AME, y = factor, xmin = lower, xmax = upper)) +
  geom_point(colour = "#939597") +
  geom_pointrange(colour = "#939597") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#F65058") +
  facet_wrap(~model, ncol = 1, scales = "free_y", strip.position = "top") +
  # facet_grid(rows = vars(model), scales = "free_y", space = "free_y")
  labs(y = NULL, x = "Average Change in Probability of Frame Usage") +
  theme(strip.background = element_rect(
    fill = "#EEEFF1",
    colour = NA)
  )

fcsts <- twofram_marg %>%
  count(model) %>%
  pull(n)

margtwoframes <- ggplotGrob(plot)
ind <- margtwoframes$layout$t[grepl("panel", margtwoframes$layout$name)]
for (i in seq_along(ind)) {
  margtwoframes$heights[[ind[i]]] = unit(fcsts[i], "null")
}
grid.newpage()
grid.draw(margtwoframes)
```

Focusing on **protest goals** first.
Table 3 analyzes protest goals compared to the reference, political protests.
Hypothesis H1a, proposing a hierarchy of social struggle that influences the use of **delegitimising** frames, is unsupported:
no significant effect is found on the use of these frames among different goals.
In other words, delegitimising frames are used similarly across all goals, refuting a social struggle hierarchy in UK newspaper coverage.
However, as each article can contain one or multiple delegitimising or legitimising frames or neither, this result has no impact on H1b , as the two models are independentl from each other.

In fact H1b, regarding a social struggle hierarchy influencing **legitimising** framing, is partially supported.
Labour protests (AME: 6%) and social-issue protests (AME: 16%) are more likely to receive legitimising coverage than political protests.
For instance, a typical labour issue, like a protest for police salaries, has a 6% higher chance of being reported with legitimising framing, and a protest against tuition fees (a social issue) is 16% more likely to be legitimised compared to political protests.
However, anti-war and police protests show no significant difference from the reference category.
H1b can this be accepted only partially:
protests with certain goals are more likely to be covered with a legitimising frame than others.
However, the specific hierarchy in the UK appears different from what has been found previously for the US context [@kilgoHierarchy2019].
Specifically, protests against police brutality have the least chance of legitimising coverage.
Then come political protests, anti-war protests, labour protests, and finally, social-issue protests with the highest probability of legitimising framing.
However, there is uncertainty in this hierarchy as political and anti-war protests show nearly identical probabilities of legitimising framing.

As explained before, the relation between violent **tactics** and less favourable coverage of the protest is well established.
Table 3 and @fig-margtwoframes confirm this connection.
Violence has a significant effect in the expected directions in both the models for legitimising and delegitimising frames.
When there is violence, the probability that a delegitimising frame is used increases by `r marg("violent protest", unit = "percent")`.
Conversely, when protesters remain peaceful, the likelihood for usage of legitimising frames increases by `r marg("violent protest", model = "legitimising", unit = "percent")`.
In other words, when protests become violent, it is a lot more likely that they will receive both more delegitimising and less legitimising coverage.
Consequently, both H2a and H2b are accepted.
Furthermore, some authors suggest that protests might use violent tactics to attract more media coverage [e.g., @boyleAdherenceProtestParadigm2012].
This might still make sense, but given the significantly more negative coverage, media exposure for a specific protest needs to be quite high for this trade-off to make sense.

The notion that the **state's response** matters for protest coverage is based on findings by @WasowSeeding2020, which suggest that in cases with violent reactions towards peaceful protest, coverage will be more legitimising.
However, the findings here show the exact opposite effect:
it appears that legitimising framing is *less* likely by `r marg("repression of peaceful p.", model = "legitimising", unit = "percent")` for events in which the state uses repressive tactics against peaceful protesters.
This means that H3 has to be rejected.
But why?

It must be noted that the effect here is not driven by *violent* state responses to peaceful protest, as was the case in the study by @WasowSeeding2020.
There were no instances of violent state response towards peaceful protest in the UK data set --- or at least they have not been coded by @ClarkMMPdata2019.
Instead the variable *state repression of peaceful protest* reflects *arrests* of peaceful protesters here.
The findings are nevertheless striking; it means that when protesters are arrested, this has a significantly negative impact on the probability of receiving coverage framed in a legitimising way, even if they remain peaceful.

On the outlet level, three factors are taken into account: *newspaper ideology*, *ideological divide* and *newspaper type*.
Despite a strong theoretical underpinning, **newspaper ideology** shows no significant effect in either of the two models in Table 3.
This means both H4a and H4b have to be rejected.
The outcome is surprising, especially since the UK case was specifically chosen because the ideological divide between outlets is more distinct than in the US or elsewhere [@kuhnPoliticsMediaBritain2007; @hallinComparingMediaSystems2004] and has been suggested to shape UK coverage of protest as well [@gavinPressTelevisionBritish2007].

Instead, one might assume that **ideological divide** might instead explain coverage.
In fact, during the manual part of the content analysis, it became apparent that some protests were supported openly by right-leaning news outlets, such as the 2000 fuel protests or protests against fox hunting legislation introduced in 2005, while left newspapers were much more critical.
So maybe ideological conflict, rather than newspaper ideology, matters for reporting?
Again, the answer is no.
Both models show no significant difference between ideological congruence and conflict between protesters and newspaper.
Both H5a and H5b are therefore rejected.
This means that while there is anecdotal evidence that left-wing media are more critical of right-wing protest and vice versa, the phenomenon is not widespread enough to be picked up by systematic statistical analysis.

Surprisingly, the *ideological ambiguity* category, which comprises very heterogeneous events ranging from protest against business practices of a company to ex-pat protests about issues abroad, has a positive effect in both models.
It is thus not straightforward to understand why they are more likely to be covered with legitimising frames *as well* as delegitimising frames.
One possible explanation is that journalists feel less constrained to voice their opinion when an issue hasn't been discussed along party lines (yet).

Moving on to the *type of newspaper*, we can see that this indicator only has an effect in the model for legitimising frames:
it is significantly less likely for tabloid news media to use legitimising frames than it is for broadsheet outlets.
Specifically, the probability that a legitimising frame is used is `r marg("tabloid newspaper", model = "legitimising", unit = "points")` percentage points lower on average when an article is published in a tabloid newspaper.
For delegitimising frames, it seems that both types of media outlets are equally likely to use delegitimising frames.
Hypothesis H6a is therefore rejected, while H6b is accepted.

Two of the tested factors were related to **timing** of events.
The model for delegitimising frames shows that as more days pass after the beginning of a protest, it becomes more likely that articles will contain a delegitimising frame.
The AME shows that the effect is one of the largest at a -`r marg("days from start", unit = "percent")`.
However, note that *days from start* was one of the two variables that were rescaled so that the minimum was 0 and the maximum was one.
The AME here, therefore, displays the change in probability when going from the minimum of 7 days before, to the maximum of 81 days after the start of a protest.
The relation is shown in @fig-dfspred and it becomes clear that it is not  a linear one.
Instead, we see the highest average probabilities that one of the delegitimising frames is used in the first week after a protest, with lower values before and after.
Average probabilities decrease for a while but increase again after about 8 weeks --- a time frame in which only a few high-profile and long-lasting protests still receive any coverage at all.
It seems then that delegitimising frames are most compelling to journalists in the direct aftermath of an event.
This confirms the expectation in H8a.

```{r}
#| label: fig-dfspred
#| fig-cap: Predicted Probabilities of Using a Delegitimising Frame Before and After the Start of a Protest (dotted red line)
days_pred <- prediction::prediction(models$model_logit_dele_mlm) %>%
  # scale back to original values
  mutate(days_since_start = scales::rescale(
    days_since_start,
    to = c(-7, 81))
  )
days_pred %>%
  ggplot(aes(x = days_since_start, y = fitted)) +
  geom_point(colour = "#939597", alpha = 0.2) +
  geom_smooth(method = "loess", formula = "y ~ x", se = FALSE, colour = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#F65058") +
  scale_x_continuous(breaks = c(-7, 0, 5, 10, 20, 30, 40, 50, 60, 70, 80)) +
  labs(y = NULL, x = NULL)
```

As was already shown in @gruberTrouble2023, the year doesn't affect the probability of delegitimising frames but does increase the probability that legitimising frames being used.^[
Although @gruberTrouble2023 uses the year in which an article was published while here it is the year of the protest event itself --- which is identical for most articles though. Note also that articles were excluded if they could not be matched with any event in the MMP data.
]
The full model, therefore, confirms this previous finding as the year has a significant positive effect on the probability that legitimising framing is used, suggesting legitimising frames become more salient over time.
Specifically, comparing the last year, 2017, in the dataset to the first year, 1992, legitimising framing has become `r marg("Year of protest (since 1992)", model = "legitimising", unit = "percent")` more likely on average.
H8b is this confirmed.

## Unpacking Frame Usage

In the previous section, the theoretical expectations towards what influences the usage of frames were tested.
This was done using grouped dependent variables signifying if one of several delegitimising or legitimising frames were used.
However, the use of this dichotomy has recently been criticised as potentially masking importan mechanisms in protest reporting [@harlowNewParadigm2023].
This section, therefore, discusses regression results for seven different models, each using a different frame as the dependent variable.
The data and model specifications for all of these models were identical to what was discussed above.
Discussion of the results follows the same structure as the previous section using Table 4 and the AME in @fig-margindframes.

```{r}
#| results: asis
indfram_marg <- readRDS("../misc/model_margins_ind.Rds")

marg_ind <- function(...) {
  marg_default(df = indfram_marg, ...)
}

dep_ind <- c("Police Violence", "Law & Order", "Cause & Grievances", "Righteous Struggle", 
"Troublemakers", "Decay of Morals", "Nuisance")

models_sel <- models[c("model_logit_1_mlm", "model_logit_2_mlm",
                       "model_logit_3_mlm", "model_logit_4_mlm",
                       "model_logit_5_mlm", "model_logit_6_mlm",
                       "model_logit_12_mlm")]
reorder <- c(2, 5:7, 3:4, 1)


if (is_latex()) {
  texreg(
    models_sel[reorder],
    #custom.model.names = paste0("frame ", c(1:6, 12)),
    custom.model.names = gsub("&", "\\&", dep_ind[reorder], fixed = TRUE),
    custom.header = list("Delegitimising" = 1:4,
                         "Legitimising" = 5:6),
    caption = "Regression Results for Individual Frames",
    custom.coef.names = c(
      "(Intercept)",
      "goal: anti-war ",
      "goal: labour protests",
      "goal: police",
      "goal: social-issue",
      "violent protest",
      "repression of peaceful p.",
      "right-wing",
      "tabloid newspaper",
      "ideological divide: conflict",
      "ideological divide: ambiguous",
      "days from start",
      "year of protest"
    ),
    groups = list("Event-Level Factors" = 2:7, "Outlet-Level Factors" = 8:11, "Time Bound Factors" = 12:13),
    digits = 3,
    stars = c(0.001, 0.01, 0.05, 0.1),
    symbol = "+",
    scalebox = 0.7,
    siunitx = TRUE,
    single.row = FALSE,
    booktabs = TRUE,
    sideways = TRUE,
    label = "tbl-regindframes",
    use.packages = FALSE
  )
} else {
  data.frame(placeholder = "In the PDF, the table is around here") %>%
    custom_df_print(caption = "Regression Results for Individual Frames")
}
```

```{r}
#| label: fig-margindframes
#| fig-cap: Average Marginal Effects for Significant Variables in Individual Frame Models
#| fig-height: 12
#| fig-width: 7
plot <- indfram_marg %>%
  ggplot(aes(x = AME, y = factor, xmin = lower, xmax = upper)) +
  geom_point(colour = "#939597") +
  geom_pointrange(colour = "#939597") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#F65058") +
  facet_wrap(~model, ncol = 1, scales = "free_y", strip.position = "top") +
  #facet_grid(rows = vars(model), scales = "free_y", space = "free_y") +
  labs(y = NULL, x = "Average Change in Probability of Frame Usage") +
  theme(strip.background = element_rect(
    fill = "#EEEFF1",
    colour = NA
  ))

fcsts <- indfram_marg %>%
  count(model) %>%
  pull(n)

# weird hack to control space between axis ticks
margindframes <- ggplotGrob(plot)
ind <- margindframes$layout$t[grepl("panel", margindframes$layout$name)]
for (i in seq_along(ind)) {
  margindframes$heights[[ind[i]]] = unit(fcsts[i], "null")
}
grid.newpage()
grid.draw(margindframes)
```

In the previous section, I concluded that none of the **goals** has a significant effect on the probability that any delegitimising frame is used.
Yet, on an individual frame level, several significant values are observed in Table 4.
The Nuisance frame is significantly more likely to be utilized in the case of labour protests than for the control category (AME: `r marg_ind("goal: labour protests", model = "Nuisance")`).

For protests against police brutality, the effects on various frames counteract each other.
It's considerably more likely that Law & Order (AME: `r marg_ind("goal: police", model = "Law & Order")`) and the Troublemakers (AME: `r marg_ind("goal: police", model = "Troublemakers")`) frames are used in the coverage of police protests.
Conversely, it is less likely that the Decay of Morals (AME: `r marg_ind("goal: police", model = "Decay of Morals", abs = FALSE)`) and/or the Nuisance (AME: `r marg_ind("goal: police", model = "Nuisance", abs = FALSE)`) frame are utilized.
This means some delegitimising frames are more probable while others are less so, causing the likelihood of any delegitimizing frame being used to remain roughly the same.

The chances of the Law & Order frame being used, revealing the positive influence of police presence, increases for protests against police brutality.
However, a higher AME is observed in the model for the antithetical frame, Police Violence (AME: `r marg_ind("goal: police", unit = "percent", model = "Police Violence")`).
That means that protests against police brutality are divisive:
they lead to more coverage highlighting claims of police brutality and providing background on them;
but coverage will also defend police actions.
They lead to more coverage emphasizing claims of police brutality while also defending police actions.

The final goal, social issues, only significantly alters the result in the model for the Law & Order frame.
Specifically, it's `r marg_ind("goal: social-issue", unit = "percent", model = "Law & Order")` less probable that the Law & Order frame is utilized for social-issue protests in comparison to political protests.

In the model for aggregated legitimizing frames, social issue and labour protests were given preference.
Here we see that this effect is solely due to the increased likelihood of *Cause & Grievances* being used (AME: `r marg_ind("goal: labour protests", model = "Cause & Grievances", abs = FALSE)`).
While the cause & grievances frame is significantly more probable (AME: r marg("goal: social-issue", model = "Cause & Grievances", abs = FALSE)), the righteous struggle model doesn't show any significant effect.

The remaining frame, police violence, neither legitimising nor delegitimising, displays the most obvious evidence that different goals influence the journalists' choice to employ a frame.
Anti-war and political protests use this frame in the same manner.
All other demands lead to significantly different uses of this frame.
Labour and social-issue protests are less likely to resort to police violence (AME: `r marg_ind("goal: labour protests", model = "Police Violence", abs = FALSE)`, `r marg_ind("goal: social-issue", model = "Police Violence", abs = FALSE)`), whereas anti police brutality protests are significantly more likely to be covered with this frame (AME: `r marg_ind("goal: police", model = "Police Violence", abs = FALSE)`).

The outcomes for individual frames corroborate the theoretical idea behind H1a-b:
the goal of a protest matters for how it's covered.
However, results are more complex than anticipated.
Not all protest events with certain types of goals are covered with frames that are more or less favourable for protesters.
Instead, different protest demands seem to steer the focus of reporting.


When it comes to protest **tactics**, the models for individual frames suggest that the positive impact on delegitimising framing was propelled by the use of the *Law & Order* (AME: `r marg_ind("violent protest", model = "Law & Order", abs = FALSE)`) and *troublemakers* (AME: `r marg_ind("violent protest", model = "Troublemakers", abs = FALSE)`) frames.
The other two delegitimising frames actually have significant negative effects, although of a lesser magnitude.
This is logical as both *law & order* and *troublemakers* are primarily centered around violence. 

*Decay of morals* and *nuisance*, on the other hand, are less likely used respectively by `r marg_ind("violent protest", model = "Decay of Morals", unit = "percent")` and `r marg_ind("violent protest", model = "Nuisance", unit = "percent")` when violence is present.
In such scenarios, journalists who wish to highlight the negative aspects of an event likely opt for a frame that directly focuses on violence.
This notion is supported by the *police violence* model, which indicates that journalists tend to exercise a healthy amount of scepticism when violence unfolds during a protest.
Although AMEs are not directly comparable across models, it's evident that the average change of probability, `r marg_ind("violent protest", model = "Police Violence", unit = "points")` percentage points, is greater than in the models for both *Law & Order* and *Troublemakers*. 
Therefore, in this aspect, the model for the use of *Police Violence* is more akin to those for delegitimising frames.

Table 4 further reveals that the negative effect for violence seen in the legitimising model was entirely driven by *cause & grievances*.
As violence occurs, journalists are less likely to follow the reasoning of protesters, a finding which basically supports what is already common knowledge.
Interestingly, the model for *righteous struggle* is the only one that shows no significant variance when violence is involved in an event.
This implies that journalists depict a cause as just regardless of violence occurrence, which seems somewhat hard to believe.
However, remember that the use of this frame is also independent of other frames' use.
In other words, even when journalists acknowledge a cause as just, the occurrence of violence still prompts them to concentrate on tactics over the event's message.

The models for individual frames present a somewhat perplexing scenario when it comes to the variable *state repression of peaceful protest*.
Despite the delegitimising frames model indicating no effect, three out of the four frames in the category do exhibit significant values.
However, the positive probabilities for *law & order* and *nuisance* seem to be offset by the negative effect for *decay of morals*.

This suggests that while it is `r marg_ind("repression of peaceful p.", model = "Law & Order", unit = "percent")` more likely that *law & order* is used, and `r marg_ind("repression of peaceful p.", model = "Nuisance", unit = "percent")` more likely that *nuisance* is used, the overall delegitimising framing remains stable when protestors are *arrested*, as *decay of morals* is used significantly less frequently (AME: `r marg_ind("violent protest", model = "Decay of Morals")`).
It appears that when arrests occur, trouble is emphasized, but as the *troublemakers* model shows no significant impact, protestors are not directly held accountable for it.

It's noteworthy that a negative effect exists for *state repression of peaceful protest* in the *decay of morals* model.
This frame underscores the risk of protestors setting a bad example for others and disrupting the general political consensus.
One potential explanation could be that journalists favor using a frame that accentuates trouble when protestors are arrested, believing that it adequately covers what they would otherwise mention about morals.
However, the presence of a significant positive effect on the usage of the *nuisance* frame contradicts this hypothesis.

In the two models for legitimising frames, the narrative from the aggregated model is echoed:
there's a significant negative effect on the likelihood of coverage using either *cause & grievances* or *righteous struggle*.
Theoretical expectations of what happens when there's *state repression of peaceful protest* fit even less well for the models of individual frames. 
The rule seems to be that getting arrested, even during peaceful demonstrations, leads to reduced probabilities of coverage that would benefit a protest.
The sole exception to this rule is the *decay of morals* frame, which is used less frequently in these instances.

On the outlet level, one of the main questions was if a *newspaper's ideology* or the distance between the newspaper's and protesters' ideology, i.e., the *ideological divide*, matters for reporting --- with the surprising result that neither of them does.
This is only slightly different for individual frames.
There is a significant negative effect in the models for *righteous struggle*:
for articles published in right-leaning outlets, the probability that this frame is used is `r marg_ind("right-wing", model = "Righteous Struggle", unit = "points")` percentage points lower compared to left-leaning outlets.
This result is little surprising as *righteous struggle* basically highlights the advantages protest has for an issue.
More surprising is the complete lack of other significant effects of this variable in the other models.
There is an effect for *law & order*, suggesting that the frame is *less* common in right-leaning outlets.
Alas, the p-value is 0.09, which is usually not considered significant.
Overall then, it seems that *newspapers' ideologies* play a small role when it comes to the framing of protest events.

As in the aggregated models, there is no support for the two hypotheses regarding the impact of an *ideological divide* between protest and reporting newspaper.
Although all of the models for delegitimising frames show p-values just over the significance threshold of $p < 0.05$.
And defying theoretical expectations again, the *police violence* frame, which highlights the responsibility of the police for clashes, is `r marg_ind("ideological divide: conflict", model = "Police Violence", unit = "percent")` more likely when there is an ideological divide.
The impression from the last section that the impact of newspaper ideology on protest coverage is incidental instead of systematic is further reinforced by these findings.

Only two of the seven models for the usage of individual frames show any significant effect for **type of newspaper**:
*righteous struggle* is used `r marg_ind("tabloid newspaper", model = "Righteous Struggle", unit = "percent")` less likely by tabloid newspaper, *police violence* is used `r marg_ind("tabloid newspaper", model = "Police Violence", unit = "percent")` more likely.
This means that **H6b** was accepted because *righteous struggle* is used more likely in quality media, while the newspaper type has no significant effect on the other legitimising frame.
All delegitimising frames are used regardless of the type of newspaper, which confirms the finding from the last section.




In terms of **temporal factors**, my findings for the aggregated frames suggest that the *days since the start* of a protest only significantly influence delegitimising frames. 
This story is repeated in the models for individual frames, providing more evidence that the theory of a *news framing cycle* [@gottliebProtestNewsFraming2015], does not hold up for legitimising framing in the UK:
instead of becoming more common after the first phase of a protest, *righteous struggle* is less likely to be used over time, while there is no effect for *cause & grievances*.

For two of the four delegitimising frames, however, the theoretical expectation is further supported:
they are less likely to be used the longer a protest is in the past.
However, there is no effect for *nuisance* and a positive effect for *decay of morals*.
This fits @gottliebProtestNewsFraming2015’s argument who suggests not that specifically delegitimising frames are used in the early stages of protest coverage, but that "conflict frames" (p. 7) are employed most often.
When conflict itself becomes less interesting, it appears that *decay of morals*, which highlights that protesters are a fringe group and set a bad example, gains in importance.

The models for individual frames also provide further evidence that legitimising frames gain in importance over the years.
Compared to 1992, the use of *cause & grievances* is `r marg_ind("Year of protest (since 1992)", model = "Cause & Grievances", unit = "percent")` more likely in 2017.
The only delegitimising frame that is influenced by a temporal shift is *law & order*, which becomes less likely to be used as time goes on (AME: `r marg_ind("Year of protest (since 1992)", model = "Law & Order", abs = FALSE)`).
Interestingly, *Police Violence*, again, aligns with the trend of the legitimising frames, as Table 4 reveals a significant positive effect for the *year* variable (AME: `r marg_ind("Year of protest (since 1992)", model = "Police Violence", abs = FALSE)`).


# Discussion and Conclusion

The analyses presented in this article provide further evidence that the importance of the *protest paradigm* for news framing of protest is more conditional than suggested several decades ago [also e.g., @kilgoHierarchy2019; @shahinProtestingParadigmComparative2016; @leeTriggeringProtestParadigm2014].
Or in other words, not all protests are treated the same and hence not all voices are heard to the same degree.
Specifically, the systematic analysis of a set of key factors that are expected to shape frame usage in the 5,639 articles in this study showed that the tactics and goals, arrests of protesters, the days between a protest and a report about it play important roles for how protests are framed in the media.
Additionally, legitimising framing has become more common in recent years.
The analysis presented in this chapter is, therefore, an important step toward understanding how different protests are treated by the media.

So which factors do explain the usage of different frames?
As it turns out, the answer depends on which frame one looks at.
For the aggregated frames, it seems that delegitimising framing were used more arbitrarily.
Their employment is significantly influenced by the protesters' goals, arrests, the media outlet, or the year of publication.
Only violence and the time since the event have a significant impact:
more violence increases the likelihood of delegitimising framing, and such framing is most likely shortly after a protest but decreases over time.
This supports the theory behind the *protest paradigm*, suggesting that barring violence, protesters have limited control over how their actions are portrayed.

However, considering only delegitimising frames doesn't provide a complete picture.
Reports on peaceful protests, particularly those advocating for labour or social issues, published in broadsheet outlets in recent years, are most likely to feature legitimising frames.
On important contribution of this article is then to show that factors shaping the usage of legitimising and delegitimising frames are not entirely antagonistic, except in the case of violence.

The second area in which is study departs from previous research is that it also considers effects on the level of individual frames.
Zooming in on this level provides some additional insights.
Importantly, while delegitimising framing seemed somewhat arbitrary, the individual frames served a clear purpose for the press:
Peaceful protests are often framed as a nuisance or reported on with the Decay of Morals frame, portraying protesters as fringe groups with immoral ideas.
Violent protests are represented as troublemakers needing to be curtailed through law and order tactics or even police violence.

A similar pattern is seen with legitimising frames:
the Cause & Grievances frame is more likely applied to labour and social-issue protests, while the Righteous Struggle frame shows no significant relation to protest goals.
Cause & Grievances is less likely used when violence occurs or when protesters are arrested and becomes more prevalent over time.

The analyses presented here directly speak to several theories:
The most robust finding is that violence predicts the usage of different frames in reports on protest.
The *protest paradigm* also predicts the measured framing reasonably well --- at least on the aggregated level and for delegitimising framing.
At least when the paradigm is defined as a default mindset when it comes to reporting about protests, which suggests that a majority of reporting uses delegitimising frames, ignoring most specific features of an event.
However, *protest paradigm* does not explain the growing prevalence of legitimising frames, which co-exist with delegitimising coverage.
While other scholars have found the *protest paradigm* to be more conditional though, the analyses here provide that mostly only legitimising framing is influenced by the tested factors.

For the theory of a *hierarchy of social struggle* presented by @kilgoHierarchy2019, the picture is mixed.
The demands of a protest *do* matter;
however, a clear hierarchy did not emerge.
The most plausible hierarchy can be observed at the level of aggregated legitimacy frames:
social-issue protests are given precedence over protests with a *labour* issue, *anti-war*, *political* protest and finally protests against police brutality.
Notably, this order is also different from what previous studies found and what had been theoretically expected [@boyleAdherenceProtestParadigm2012; @boyleNewspapersProtestExamination2005; @boyleInfluenceLevelDeviance2004].
Furthermore, some of these differences, are not statistically significant.
The results for individual frames suggest that rather than following one clear hierarchy, different protest demands seem to lead to a different focus in reporting.
Generally though, except for the differences between goals observed in the application of the *police violence* frame, the usage of frames between demands appears rather similar, with only one or two demands following a different trajectory in each frame.
The idea that some protest goals are treated more favourably by the media is, therefore, rejected for the UK case.

The theory presented by @WasowSeeding2020 that *state repression of peaceful protest* leads to more legitimising coverage of said protest could also not be confirmed.
In part, this was probably due to a lack of available cases, as not a single case of police violence against peaceful protesters was present for the UK in the MMP data.
The employed proxy for state repression, police arrests, causes an effect in the opposite direction:
legitimising framing was less likely to be used.
Getting arrested, even while peaceful, diminishes the chances for protesters to get their message on the media agenda.
The *police violence* frame is also more likely to be used in these cases;
however, it is not at all clear if media reports using this frame aide protesters' causes.

The idea that the *organisation* level, which comprises the larger context of the routinised activities such as the goals, policies and structure of a news organisation [@shoemakerMediatingMessage21st2014] influences the framing of protest events was also mostly rejected by the analysis here.
The *ideology* of a news outlet matters only for the *righteous struggle* frame, while an ideological divide between protesters and outlet matters only for the *police violence* frame (AME = `r marg_ind("ideological divide: conflict", model = "Police Violence", abs = FALSE)`).
Overall then, it seems that *newspaper ideology* plays a small role when it comes to framing of protest events in the UK.

The difference between tabloid and broadsheet news outlets is surprisingly also rather small according to the analysis presented here.
Only two frames are used differently between the newspaper types:
the *police violence* frame, which attributes responsibility for clashes during protests to police is more common in tabloid newspapers;
the legitimising *righteous struggle* frame, which highlights the positive effect of protests for their respective issues, is more common in broadsheets news.
The result is surprising as previous studies assumed that the stark divide between newspaper titles with different ideologies and of different types cater to different audiences and would hence prefer different frames [@cammaertsSymbolicDamage2013; @gavinPressureGroupDirect2010].
While this might certainly be true for individual protests, and has been found to be the case occasionally during manual coding in this study, the systematic analysis of all protests events and coverage in the UK did not replicate these findings.

Finally, the analyses assed the existence of a *news framing cycle*:
just as posed by @gottliebProtestNewsFraming2015, coverage of protest begins with a focus on conflict --- especially using the *troublemakers* and *law & order* frames.
Unlike what was found by @gottliebProtestNewsFraming2015, though, these frames are not replaced with legitimising ones later on.
In fact, the *righteous struggle* frame is also used less frequently after some time following a protest.
Rather, it appears that the conflict frames are replaced by the *decay of morals* frame as time passes after the start of an event.
The different finding here might be due to the main force @gottliebProtestNewsFraming2015 expects to drive the *news framing cycle*:
growing participation.
Yet, increased participation after the first day of an event is only common for few high-profile events that stretch over days and months --- like the Occupy Wall Street movement for which @gottliebProtestNewsFraming2015 developed the theory.
Nevertheless, a variable that includes days after a protest started seems to be a worthwhile inclusion for studies that test factors of influence for protest reporting.

This study is not without limitations.
Firstly, it is limited in that it only focuses on newspapers and only addresses the state of protest coverage in the UK.
Some interrogated factors for protest coverage might be more or less influential in other forms of media and in other contexts.

Secondly, the employed *Mass Mobilization Project*'s [@ClarkMMPdata2019] dataset on protest events introduces endogeneity concerns to the study as the data was collected from media reports.
As the dependent variable is media reporting patterns, the fact that some independent variables are sourced from this reporting might create a feedback loop complicating our ability to separate cause and effect.
This potentially affects the measurement of the variables *Protest Goal*, *Protester Violence* and *State Repression*.
However, this is risk is currently unavoidable as independent data does not exist.
All the exiting databases that I could have used alternatively suffer from the same problem, while media-independent information about events is scarce and might be equally biased:
@mccarthyElectronicPrintMedia1999, for example, use police records, which arguably suffer from other biases.
To set this in context, however, research using media-independent information about events has shown that publishing outright false information about an event is relatively rare [@mccarthyElectronicPrintMedia1999].
Reports rather omit information or misrepresent what happened during an event through framing [@earlUseNewspaperData2004] --- like emphasising, for example, violence even when violence was only a very small aspect of what happened.

\newpage
# References {.unnumbered}

::: {#refs}
:::

\newpage
# Appendix {#sec-eventdata .unnumbered}

```{r article, child = "app-eventdata.qmd"}
```
